---
title: "Results"
author: "Gareth Edwards"
date: "12/11/2020"
output:
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = 'hide')
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.keep = 'all')

source("R/cwac_functions.r")
library(knitr)
library(vegan)

# load relevant data
load('data/robertsDB.RData')
load('data/barberspan_counts.RData')
```

# Results

This study takes into account 3 objectives. These 3 objectives, in short, are to perform exploratory data analysis on the CWAC dataset, develop a state space time series model to model the bird counts over time (both individually and collectively) and, finally, to formulate relevant bioindicators for wetland conservation decision makers.

## Exploratory Data Analysis (EDA)

The CWAC dataset contains bird counts from various wetlands across South Africa. The counts normally occur biannually, once during the summer and once during the winter. Only the Barberspan wetland has been used for this report. There are bird counts ranging from the year 1993 all the way to 2018 in the Barberspan records. The original dataset omitted the years where no counters reported any counts. These years were inserted into the dataset with NA values representing the bird counts for those years. This was done for the sake of completeness and for the benefit of the state space models. 

The resident and migratory species have been analysed separately to account for their differences in migratory patterns. These patterns may present themselves in the data. For example, we would likely see greater fluctuation in the migratory bird counts over the years compared to the resident species, whose counts would stay relatively stable throughout the year.

Below (probably will put in an appendix) we see boxplots for each bird species in the Barberspan wetland. 

```{r EDA, fig.cap="Resident species boxplots"} 
# make captions work

birdcnts <- counts[,1:106]

# separate into resident and migrant
resid <- robertsDB %>%
  filter(Migrant=="n") %>%
  select(id)

migid <- robertsDB %>%
  filter(Migrant=="y") %>%
  select(id)

residents <- birdcnts[,resid$id]
migrants <- birdcnts[,migid$id]

# plot a grid of boxplots for the resident species
par(mfrow = c(6,3))
par(mar=c(1,2,2,1))
for(i in 1:18){
  boxplot(residents[,i], ylab = "Counts", main = colnames(residents)[i])
}

for(i in 19:36){
  boxplot(residents[,i], ylab = "Counts", main = colnames(residents)[i])
}

for(i in 37:54){
  boxplot(residents[,i], ylab = "Counts", main = colnames(residents)[i])
}

for(i in 55:72){
  boxplot(residents[,i], ylab = "Counts", main = colnames(residents)[i])
}

resnas <- as.vector(apply(residents, 2, function(x){
  return(sum(!is.na(x)))
}))
rareres <- sum(resnas<6)

```

```{r EDA2}

# plot a grid of boxplots for the resident species
par(mfrow = c(6,3))
par(mar=c(1,2,2,1))
for(i in 1:17){
  boxplot(migrants[,i], ylab = "Counts", main = colnames(migrants)[i])
}

for(i in 18:34){
  boxplot(migrants[,i], ylab = "Counts", main = colnames(migrants)[i])
}

mignas <- as.vector(apply(migrants, 2, function(x){
  return(sum(!is.na(x)))
}))

raremig <- sum(mignas<6)

```

There are a considerable amount of species that contain fewer than 6 counts. These species are considered rare in this study. In total there are `r rareres+raremig` rare species in Barberspan. `r rareres` rare resident species and `r raremig` rare migrant species. These species may not be certified rare species but are categorised this way in this study based on the data alone. The fact that there are so many species with few counts could be attributed to many factors that will be discussed in later sections. Below is a bar plot showing the number of counts each species in Barberspan contains.


```{r EDA3}
# produce a histgram of NA values for each species in Barberspan
nas <- as.vector(apply(birdcnts, 2, function(x){
  return(sum(is.na(x)))
}))

barplot(nas)

```

Further exploration is conducted with each year as a focal point. Below we see a bar plot with total counts in each year. There are some years where there are no counts for any species. The time periods with no bird counts are as follows: The winter of 1995, the winter of 1996 till the winter of 1998, the winter of 2007, the winter of 2010 till the winter of 2014, the summer of 2016 till the winter of 2017 and the winter of 2018. In total there are 21 different time periods that contain no counts. This is largely due to the fact that years were inserted into the original dataset to create a continuous time period from 1993 till 2018.

```{r EDA4}

totalcnts <- apply(birdcnts, 1, sum, na.rm = T)
barplot(totalcnts)

```


```{r EDA5}
# 2 boxplots showing the resident species and the migrant species in Barberspan

totrescnts <- apply(residents, 1, sum, na.rm = T)
totmigcnts <- apply(migrants, 1, sum, na.rm = T)

par(mfrow=c(1,2))
boxplot(totrescnts, main = "Resident species")
boxplot(totmigcnts, main = "Migratory species")

```

The Boxplot above shows the distribution of the resident and migratoy species in Barberspan. There are significantly more resident species than migratory species in Barberspan. There may be various reasons for this that will be discussed later. 

## State Space Time Series Model

### Model Selection

State space time series models were used to model the bird counts across the years and seasons. Three different models were considered:

1. A model for residents.
2. A model for migrants.
3. A model for rare species.

The migrant and resident models were applied to both residents, migrants and rare species. This was to monitor the DIC to identify which model worked best and if, perhaps, it was beneficial to use one model for all the species. The rare species model was only applied to rare species as it wasn't possible to apply it to the residents and migrants. Below is shown a table with the various models applied to the various types of species and their corresponding DIC score. Deviance information criterion (DIC) is commonly used for Bayesian models that produce MCMC output. DIC is the AIC equivalent in a frequentist context. DIC is similar to AIC in that it takes into account how well the model fits the data and the model complexity and finds a balance between the two. The difference with DIC is that it takes prior information into account whereas AIC doesn't (ref: DIC_theroy.pdf).

```{r dicTable}

# run resident model on all resident species
res.jags <- list()
com.res <- residents[,resnas>6]

for(i in 1:length(com.res)){
  res.jags[[i]] <- jags_com(as.data.frame(cbind(com.res[,i],
                                          'season' = counts$season,
                                          'startDate' = counts$startDate)))
}

# run migrant model on all resident species
mig.jags <- list()

for(i in 1:length(com.res)){
  mig.jags[[i]] <- jags_analysis(as.data.frame(cbind(com.res[,i],
                                          'season' = counts$season,
                                          'startDate' = counts$startDate)))
}

# get average DIC for both models
dics <- lapply(res.jags, function(x){
  return(x$DIC)
})
rres.dics <- mean(unlist(dics))

migdics <- lapply(mig.jags, function(x){
  return(x$DIC)
})
mres.dics <- mean(unlist(migdics))

# run resident model on all migrants
mig2.jags <- list()
com.mig <- migrants[,mignas>6]

for(i in 1:length(com.mig)){
  mig2.jags[[i]] <- jags_com(as.data.frame(cbind(com.mig[,i],
                                          'season' = counts$season,
                                          'startDate' = counts$startDate)))
}

# run migrant model on all migrants
mm.jags <- list()

for(i in 1:length(com.mig)){
  mm.jags[[i]] <- jags_analysis(as.data.frame(cbind(com.mig[,i],
                                          'season' = counts$season,
                                          'startDate' = counts$startDate)))
}

# obtain average DIC for both models
rmig.jags <- lapply(mig2.jags, function(x){
  return(x$DIC)
})
rmig.jags <- mean(unlist(rmig.jags))

mmig.jags <- lapply(mm.jags, function(x){
  return(x$DIC)
})
mmig.jags <- mean(unlist(mmig.jags))

# create table of DIC values
mean.dics <- c(rres.dics, mres.dics, rmig.jags, mmig.jags)
mod.name <- c("Resident Model", "Migrant Model", "Resident Model", "Migrant Model")
spec.type <- c("Residents", "Residents", "Migrant", "Migrants")

jag.tab <- cbind(mod.name, spec.type, round(as.numeric(mean.dics), 2))
colnames(jag.tab) <- c("Model Type", "Species Type", "Mean DIC")

kable(jag.tab, digits = 2,
      caption = "DIC values corresponding to model type on species type")

# obtain all rare species

# run migrant model on rare species

# run resident model on rare species

# run rare species on rare species



```




















